---
title: "numeric_eda"
author: "Emily Potts"
date: "2022-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}
weather_df =  
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, everything())
```

# group_by()
```{r}
weather_df %>% 
    group_by(name, month)
```


# summarize()

let's group then summarize!

## counting
```{r}
weather_df %>% 
    group_by(name, month) %>% 
  summarize(
    n_obs = n()
  )
```

counting other ways
```{r}
weather_df %>%
  count(month, name = "n_obs")
```

don't use table()
```{r, eval = FALSE}
weather_df %>%
  pull(month) %>% 
  table
```



```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(
    n_obs = n(),
    n_dist = n_distinct(date)
  )
```


# a brief aside... 2x2 tables
tidy
```{r}
weather_df %>% 
  mutate(
    cold = case_when(
      tmax <  5 ~ "cold",
      tmax >= 5 ~ "not_cold",
      TRUE      ~ ""
  )) %>% 
  filter(name != "Waikiki_HA") %>% 
  group_by(name, cold) %>% 
  summarize(count = n())
```

non-tidy
```{r}
weather_df %>% 
  mutate(cold = case_when(
    tmax <  5 ~ "cold",
    tmax >= 5 ~ "not_cold",
    TRUE     ~ ""
  )) %>% 
  filter(name != "Waikiki_HA") %>% 
  janitor::tabyl(name, cold)
```

# general summaries
```{r}
weather_df %>%
  group_by(month) %>%
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    mean_prec = mean(prcp, na.rm = TRUE),
    median_tmax = median(tmax, na.rm = TRUE),
    sd_tmax = sd(tmax, na.rm = TRUE)
  )
```

group by more than one variable
```{r}
weather_df %>%
  group_by(name, month) %>%
  summarize(
    mean_tmax = mean(tmax),
    median_tmax = median(tmax))
```

summarize multiple columns using the same summary, the across function is helpful
```{r}
weather_df %>%
  group_by(name, month) %>%
  summarize(across(tmin:prcp, mean))

```

create a plot based on the monthly summary
```{r}
weather_df %>%
  group_by(name, month) %>%
  summarize(mean_tmax = mean(tmax)) %>%
  ggplot(aes(x = month, y = mean_tmax, color = name)) + 
    geom_point() + geom_line() + 
    theme(legend.position = "bottom")
```

The results of group_by() and summarize() are generally tidy, but presenting reader-friendly results for this kind of exploratory analysis often benefits from some un-tidying. For example, the table below shows month-by-month average max temperatures in a more human-readable format
```{r}
weather_df %>%
  group_by(name, month) %>%
  summarize(mean_tmax = mean(tmax)) %>% 
  pivot_wider(
    names_from = name,
    values_from = mean_tmax) %>% 
  knitr::kable(digits = 1)
```







