---
title: "visualization2_with_ggplot2"
author: "Emily Potts"
date: "2022-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)
library(ggridges)
```

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())
weather_df
```


## scatterplot, but better this time:
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5)
```


labels:
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package"
  )
```

scales:
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package") + 
  scale_x_continuous(
    breaks = c(-15, 0, 15), 
    labels = c("-15º C", "0", "15"))
```
Aesthetic mappings determine which variables map to which plot attributes. These mappings have reasonable default behaviors, but can be modified through scales.

For example, you’ll occasionally want control over the location and specification of tick marks on the X or Y axis. These can be manipulated in scale_x_* and scale_y_* where * depends on the type of variable mapped to the x and y aesthetics (i.e. continuous vs discrete).

scale options:
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package") + 
  scale_x_continuous(
    breaks = c(-15, 0, 15), 
    labels = c("-15ºC", "0", "15"),
    limits = c(-20, 30)) + 
  scale_y_continuous(
    trans = "sqrt", 
    position = "right")
```

different version - scale colors:
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package") + 
  scale_color_hue(name = "Location", h = c(100, 300))
```

scale colors with viridis package:
```{r}
ggp_temp_plot = 
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package"
  ) + 
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE
  )

ggp_temp_plot
```
We used discrete = TRUE because the color aesthetic is mapped to a discrete variable. In other cases (for example, when color mapped to prcp) you can omit this argument to get a continuous color gradient. The viridis::scale_fill_viridis() function is appropriate for the fill aesthetic used in histograms, density plots, and elsewhere.

## themes

Themes are used to modify non-data elements of a plot – they don’t change mappings or how data are render, but control things like background color and location of the the legend. Using themes can help with general plot appearance.

For example, I frequently change is the legend position. By default this is on the right of the graphic, but I like to shift it to the bottom to ensure the graphic takes up the available left-to-right space.
```{r}
ggp_temp_plot + 
  theme(legend.position = "bottom")
```

Quick tip: legend.position = "none" will remove the legend. This is helpful when multiple plots use the same color scheme or when the legend is obnoxious for some other reason.

By default this is theme_gray; here’s theme_bw():
```{r}
ggp_temp_plot + 
  theme_bw() + 
  theme(legend.position = "bottom")
```

theme_classic():
```{r}
ggp_temp_plot + 
  theme_classic() + 
  theme(legend.position = "bottom")
```


here’s the Excel 2003 theme from ggthemes:
```{r}
ggp_temp_plot + 
  ggthemes::theme_excel() + 
  theme(legend.position = "bottom")
```

The ordering of theme_bw() and theme() matters – theme() changes a particular element of the plot’s current “theme”. If you call theme to change the some element and then theme_bw(), the changes introduced by theme() are overwritten by theme_bw().


## setting options
In addition to figure sizing, I include a few other figure preferences in global options declared at the outset of each .Rmd file (this code chunk just gets copy-and-pasted to the beginning of every new file).

```{r}
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## data argument in geom_*

We’ve seen that where an aesthetic gets mapped to a variable matters – setting aes(color = name) in ggplot can yield different results than the same setting in geom_point(). This arises from the way that ggplot objects inherit aesthetic mappings, and it turns out there’s a similar thing with the data used to make a plot.

In a contrived example, we can split weather_df into separate datasets for Central Park and Waikiki. Then we use one in the ggplot() call and another in geom_line():
```{r}
central_park = 
  weather_df %>% 
  filter(name == "CentralPark_NY")

waikiki = 
  weather_df %>% 
  filter(name == "Waikiki_HA")

ggplot(data = waikiki, aes(x = date, y = tmax, color = name)) + 
  geom_point() + 
  geom_line(data = central_park)
```

## patchwork package

We’ve seen facetting as an approach to create the “same plot” for several levels of a categorical variable, and this can get you pretty far. Sometimes, though, you want to show two or three fundamentally different plots in the same graphic: you may want to juxtapose a scatterplot and a boxplot, or show scatterplots illustrating relationships between different variables. In this case, a solution is to create each of the panels you want separately and combine panels using tools in the patchwork package:
```{r}
tmax_tmin_p = 
  weather_df %>% 
  ggplot(aes(x = tmax, y = tmin, color = name)) + 
  geom_point(alpha = .5) +
  theme(legend.position = "none")

prcp_dens_p = 
  weather_df %>% 
  filter(prcp > 0) %>% 
  ggplot(aes(x = prcp, fill = name)) + 
  geom_density(alpha = .5) + 
  theme(legend.position = "none")

tmax_date_p = 
  weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) + 
  theme(legend.position = "bottom")

(tmax_tmin_p + prcp_dens_p) / tmax_date_p
```

## Data manipulation

reordering name by hand:
```{r}
weather_df %>%
  mutate(name = forcats::fct_relevel(name, c("Waikiki_HA", "CentralPark_NY", "Waterhole_WA"))) %>% 
  ggplot(aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), color = "blue", alpha = .5) + 
  theme(legend.position = "bottom")
```


A second example reorders name according to tmax values in each name:
```{r}
weather_df %>%
  mutate(name = forcats::fct_reorder(name, tmax)) %>% 
  ggplot(aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), color = "blue", alpha = .5) + 
  theme(legend.position = "bottom")
```


A more difficult situation relates to data tidyiness. Suppose I wanted to create a three-panel plot showing densities for tmax and tmin within each location. More concretely, I want to be able to facet panels across the name variable, and create separate densities for tmax and tmin in each panel. Unfortunately, weather_df isn’t organized in a way that makes this easy.

One solution would recognize that tmax and tmin are separate observation types of a shared temperature variable. With this understanding, it’s possible to tidy the weather_df and make the plot directly:
```{r}
weather_df %>%
  select(name, tmax, tmin) %>% 
  pivot_longer(
    tmax:tmin,
    names_to = "observation", 
    values_to = "temp") %>% 
  ggplot(aes(x = temp, fill = observation)) +
  geom_density(alpha = .5) + 
  facet_grid(~name) + 
  viridis::scale_fill_viridis(discrete = TRUE)
```


# pulse data example
```{r}
pulse_data = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") %>%
  janitor::clean_names() %>%
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") %>%
  select(id, visit, everything()) %>%
  mutate(
    visit = recode(visit, "bl" = "00m"),
    visit = factor(visit, levels = str_c(c("00", "01", "06", "12"), "m"))) %>%
  arrange(id, visit)

ggplot(pulse_data, aes(x = visit, y = bdi)) + 
  geom_boxplot()
```

# FAS data example
```{r}
pup_data = 
  read_csv("./data/FAS_pups.csv", col_types = "ciiiii") %>%
  janitor::clean_names() %>%
  mutate(sex = recode(sex, `1` = "male", `2` = "female")) 

litter_data = 
  read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>%
  separate(group, into = c("dose", "day_of_tx"), sep = 3)

fas_data = left_join(pup_data, litter_data, by = "litter_number") 

fas_data %>% 
  select(sex, dose, day_of_tx, pd_ears:pd_walk) %>% 
  pivot_longer(
    pd_ears:pd_walk,
    names_to = "outcome", 
    values_to = "pn_day") %>% 
  drop_na() %>% 
  mutate(outcome = forcats::fct_reorder(outcome, pn_day, median)) %>% 
  ggplot(aes(x = dose, y = pn_day)) + 
  geom_violin() + 
  facet_grid(day_of_tx ~ outcome)
```

