---
title: "visualization_with_ggplot2"
author: "Emily Potts"
date: "2022-09-29"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
library(tidyverse)
library(ggridges)
```

This code gets the data from the course website
```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())

weather_df
```


# Basic scatterplot
```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + geom_point()
```

Make the same scatterplot, but different

```{r}
weather_df %>%
  drop_na() %>%
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()
```

Make the same scatterplot, but different again
```{r}
weather_scatterplot =
  weather_df %>%
  drop_na() %>%
  ggplot(aes(x = tmin, y = tmax))

weather_scatterplot + geom_point()
```

## fancy
```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = 0.3) + 
  geom_smooth(se = FALSE) 
```

fancy with separate panels
```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = 0.3) + 
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
```

spicing it up lol (seasonal trends)
```{r}
weather_df %>%
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point(aes(size = prcp), alpha = .5) +
  geom_smooth(se = FALSE)
```

seasonal trends with panels
```{r}
weather_df %>%
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point(aes(size = prcp), alpha = .5) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
```

hex plot
```{r}
ggplot(weather_df, aes(x = tmax, y = tmin)) + 
  geom_hex()
```

## Univariate plots
Histograms, barplots, boxplots, and violins...

histogram
```{r}
ggplot(weather_df, aes(x = tmax)) + 
  geom_histogram()
```

```{r}
ggplot(weather_df, aes(x = tmax, fill = name)) + 
  geom_histogram(position = "dodge", binwidth = 2)
```

```{r}
ggplot(weather_df, aes(x = tmax, fill = name)) + 
  geom_density(alpha = .4, adjust = .5, color = "blue")
```

boxplots
```{r}
ggplot(weather_df, aes(x = name, y = tmax)) + geom_boxplot()
```

```{r}
ggplot(weather_df, aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), alpha = .5) + 
  stat_summary(fun = "median", color = "blue")
```

ridge plot (need ggridges package)
```{r}
ggplot(weather_df, aes(x = tmax, y = name)) + 
  geom_density_ridges(scale = .85)

```

# Saving and embedding plots
```{r}
weather_scatterplot =
weather_df %>%
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point(aes(size = prcp), alpha = .5) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)

ggsave("weather_scatterplot.pdf", weather_scatterplot, width = 8, height = 5)
```
(can add pathname in front of "weather_scatterplot.pdf" to exisiting folder)

```{r}
weather_scatterplot
```





